---
title: "Final_Project_Code"
author: "Sami Gibson"
date: "`r Sys.Date()`"
format:
  html:
    toc: true # use this to display a table of contents
execute:
  message: false # use this to make sure messages don't show up
  warning: false # use this to make sure warnings don't show upl
---
# Set-up
```{r setup}
suppressPackageStartupMessages({
  library(here) # reads in here package
  library(dplyr)   # reads in dplyr package 
  library(forcats)  # reads in forcats package
  library(janitor) # reads in janitor package
  library(showtext) # reads in showtext package
  library(ggplot2) # reads in ggplot2 package
  library(tidyverse) # reads in tidyverse package
  library(broom) # reads in broom package
  library(performance) # reads in performance package
  library(DHARMa) # reads in DHARMa package
  library(tibble)
  library(knitr)
  library(ggeffects)
  })
# Stores the kelp data as an object called kelp
kelp <- read.csv(here('data',"Annual_Kelp_All_Years_20250903.csv")) 
# 
mopl <- read.csv(here('data',"MOPL_nest-site_selection_Duchardt_et_al._2019.csv")) 
```

[Here is my GitHub Repository](https://github.com/samantha-gibson05/ENVS-193DS_winter-2026_final.git)

# Problem 1. Research writing

You’re working on a research team trying to understand how environmental variables influence wetland flooding area (measured in $m^2$) in the San Joaquin River Delta. Your co-worker runs some analyses and writes up a report, giving it to you to review. In part 1 of the results section of the report, your co-worker has written: *We retained the null hypothesis that precipitation doesn’t predict flooded wetland area (p = 0.11).*

In part 2 of the results section of the report, your co-worker has written: *We retained the null hypothesis that median flooded wetland area does not differ across water year classification (wet, above normal, below normal, dry, critical drought) (p = 0.12).*

It’s great that this report is coming together, but you think they can improve on what they’ve written and make it more understandable to a non-statistical audience.

## a. Transparent statistical methods

What statistical tests did your co-worker use?

Clearly connect the test to the part that you are addressing (e.g. “In part 1, they used _______. In part 2, they used _______.”).

In **part 1**, they most likely used a simple linear regression ($H_0$:  the predictor variable does not predict the response; $H_A$: the predictor variable does predict the response) because the statement asks whether precipitation predicts flooded wetland area, which is a continuous response variable. In **part 2**, they most likely used a Kruskal–Wallis rank-sum test ($H_0$: sample 1 median = sample 2 median = sample 3 median ....; $H_A$: at least 2 sample medians are different form each other), because the statement compares median flooded wetland area across more than two groups of water year classification (wet, above normal, below normal, dry, critical drought).

## b. Figure needed

The test in part 2 seems familiar to you, but you think your coworker could provide additional context using a visualization for their test.

Describe one figure that your coworker could make to accompany the statement in part 2.

Be specific about what variables should be on the x-axis, the y-axis, and the visual components of the plot as they relate to the test in *addition* to naming the type of plot they should make.

A good figure for part 2 would be a **boxplot** with **jittered points**. The **x-axis** would be water year classification (wet, above normal, below normal, dry, critical drought). The **y-axis** would be flooded wetland area ($m^2$). **Visual components** of this plot should include: one boxplot for each water year class showing the median and interquartile range, with individual observations overlaid as jittered points so the reader can see spread, overlap, and sample size across categories. This would match the **Kruskal–Wallis** test well because it helps visualize how the distributions and medians compare across groups which is what the test aimns to discover.

## c. Suggestions for rewriting

In 1-3 sentences, write *new research statements* to include relevant components from parts a-b and a full test summary in parentheses to be transparent about the statistical method.

Be sure that your rewritten statements clearly delineate the *biological narrative* from the *statistical summary*. See lecture and workshop for examples of writing.

Note that your co-worker didn’t include any information about the test statistic, distribution, etc., and that you only know the p-value. For any part that you do not know, list that part with text. For example, you could write something like: “… $\alpha$**=significance level**…”

**Part 1**: Flooded wetland area did not show clear evidence of a relationship with precipitation in this analysis. The statistical test did not provide sufficient evidence to reject the null hypothesis of no association between precipitation and flooded wetland area (Linear regression, F(degrees freedom (numerator), degrees freedom (denominator)) =  test-statistic (F), $R^2$ = R-squared, p = 0.11, $\alpha$ = significance level).


**Part 2**: Median flooded wetland area did not differ significantly among water year classifications (wet, above normal, below normal, dry, critical drought). The statistical test did not provide sufficient evidence to reject the null hypothesis that median flooded wetland area was the same across wet, above normal, below normal, dry, and critical drought years (Kruskal–Wallis test, $\chi^2$(degrees of freedom) = test-statistic (H), p = 0.12, $\alpha$ = significance level)

## d. Interpretation

Based on your co-worker’s analyses, flooded wetland area isn’t affected by precipitation or by the classification of the water year.

In 1-2 sentences, outline:

- one additional variable that could influence the flooded area of a wetland,
- what *type* of variable it is, and
- *why* that variable might influence wetland flooding.

One additional variable that could influence flooded wetland area is **water-control management intensity**, such as whether managers actively impound or deliver water to wetlands. This would likely be a categorical explanatory variable, and it could influence flooding because the paper emphasizes that wetland extent in this system is shaped not only by precipitation but also by management actions and infrastructure that control water delivery and retention.

# Problem 2. Data visualization

Navigate to the SBC LTER’s data catalog. Find and download the dataset on giant kelp fronds in the Santa Barbara Channel to your `data` folder. (Hint: the page on the data catalog is called “KFCD abundance and size of giant kelp”.)

In your set up chunk, read in the data as an object called `kelp`.

## a. Cleaning and summarizing

Create an object called `kelp_clean` from `kelp`.

To clean and summarize the data frame, use the functions in the following code chunk. *Only use these functions*.

The functions are out of order. Put the functions in the right order and use the pipe operator to get the data frame you need.

Annotate each function to demonstrate your understanding of its use.

When you are done with all your cleaning steps:

- display 5 rows from `kelp_clean` using `slice_sample()` and
- the structure using `str()`

```{r}
kelp_clean <- kelp |> # uses raw data 
  # standardizes column names to lowercase and replaces spaces with underscores
  clean_names() |>  
  # keep only the three focal sites: IVEE, NAPL, CARP 
  filter(site %in% c("IVEE", "NAPL", "CARP")) |>  
  # remove observations with specified values for each date, transect, and quad
  filter(!(date == "2000-09-22" & transect == "6" & quad == "40")) |>
  # create column 'cite_name' with appropriate site names associated site codes
  mutate(site_name = case_when(site == "NAPL" ~ "Naples",
                               site == "IVEE" ~ "Isla Vista",
                               site == "CARP" ~ "Carpinteria")) |>
  mutate(site_name = as_factor(site_name), # converts `site_name` to a factor
         # sets the order of the factor 'site_name' for plotting etc.
         site_name = fct_relevel(site_name,"Naples", "Isla Vista", "Carpinteria")) |>
  group_by(site_name, year) |> # groups data by variables 'site_name' and 'year'
  # creates column 'mean_fronds' with mean fronds within each site-year group
  summarize(mean_fronds = mean(fronds)) |>
  ungroup()  # removes grouping from cleaned data set output

kelp_clean |> slice_sample(n = 5) # displays 5 rows from cleaned data set

str(kelp_clean) # displays structure of cleaned data set

```

## b. Understanding the data

The code chunk above includes a line that filters out frond observations from the date `2000-09-22` in transect 6 in quad 40.

In 1-2 sentences, describe why. What is the number for the value of fronds in that “quad” in that “transect”? What does that value represent? Reference the specific place you found information on what that value represents (e.g. in the Summary tab of the data page).

The observation from 2000-09-22 in transect 6, quad 40 is filtered out because the recorded value is -99999 which is a placeholder value rather than a real biological measurement. In the "Files and Data Links" tab the table row "Others" lists the value -99999 as a "Missing Value Code", meaning that the number of kelp fronds was not actually measured or recorded for that quadrat, so it should be removed before calculating summary statistics.

## c. Visualize the data

Recreate the visualization

The specific aesthetic components you need to recreate are:

- the geometries (there are two)
- the text (all axis and legend text, and the title)
- different, custom colors for each site
- the title position
- the panel (a grid without the border)
- use of a custom font (that is consistent across all text in the figure)

```{r}
# xxxx don't forget to comment
font_add_google("Lato", "lato")
showtext_auto()

ggplot(kelp_clean, aes(x = year, y = mean_fronds, color = site_name)) +
  geom_line(linewidth = 0.6) +                 
  geom_point(size = 2.6) +                   
  scale_color_manual(
    values = c(
      "Naples" = "cyan4",
      "Isla Vista" = "goldenrod4",
      "Carpinteria" = "grey60"),
    name = "Site") +
  labs(
    title = "Sites vary in mean kelp fronds per year",
    x = "Year",
    y = "Mean kelp fronds per year"
  ) +
  scale_x_continuous(breaks = seq(2000, 2025, by = 5)) +
  theme_minimal(base_family = "lato") +
  theme(
    plot.title = element_text(hjust = 0.5),  # centered title
    legend.position = "right",                          # legend outside right
    panel.border = element_blank(),                     # no panel border
    text = element_text(family = "lato"))
```


## d. Interpretation

*Be specific* with the visual components of the plot to support your characterization.

In 1-2 sentences each, describe:

- the site with the most variable mean kelp fronds per year

**Isla Vista** shows the most variability in mean kelp fronds per year. In the plot, the Isla Vista line and points (colored in brown) fluctuate widely across years, ranging from very low values near 4 fronds to peaks close to 40 fronds which produces large vertical changes between adjacent years.

- the site with the least variable mean kelp fronds per year

**Carpinteria** shows the least variability in mean kelp fronds per year. The Carpinteria line and points (colored in grey) stay within a relatively narrow vertical range and are generally between about 3 and 17 fronds with more gradually changes across years compared with the other sites.


# Problem 3. Data analysis

```{r}
# figure out how to exclude
mopl_explore <- mopl |>
  clean_names() |>
  mutate(nest_use = if_else(id == "used", 1, 0))
mopl_mod <- glm(nest_use ~ edgedist + vo_nest,
               data = mopl_explore,
               family = "binomial")
plot(simulateResiduals(mopl_mod))
summary(mopl_mod)
```

## a. Response variable

In 1-2 sentences, explain what the 1s and 0s mean in this data set biologically.

In this dataset, the response variable is nesting habitat (nest site use vs. non-use) where	a 1 would represent a site that **was used** by a Mountain Plover for nesting, and a 0 would represent an available but **unused** site. In the raw dataset, this information comes from the ID column, which is coded as "used" and "available".

## b. Purpose of study

In this problem, you will be examining the potential effects of *visual obstruction* and *distance to prairie dog colony edge* on the probability of a nest site being occupied by a Mountain Plover.

In 2-3 sentences for each predictor, describe:

- what kind of variable it is (i.e. continuous, discrete, categorical, etc.),
- how it could possibly influence the probability of Mountain Plover nest site use (i.e. as visual obstruction increases, the probability of Mountain Plover nest site use would ______________ because…), and
- where in the paper you found this information (the specific section, subsection, and paragraph)

**Predictor 1: Visual Obstruction** (`vo_nest`)
This is **continuous** predictor because it is the "height at which the vegetation is sufficiently dense to obscure a 2.5-cm wide pole when viewed from a 1-m height." As visual obstruction increases, the probability of Mountain Plover nest-site use would decrease, because plovers generally prefer nesting habitat with short, sparse vegetation and avoid areas where vegetation is dense enough to obstruct visibility. Additionally, the article states that, "plovers avoided nesting in areas with a maximum vegetation height >11 cm, corresponding with an average visual obstruction of >5 cm" (Discussion Section Paragraph 4).

**Predictor 2: Distance to Prairie Dog Colony Edge** (`edgedist`)

This is a **continuous** predictor, because it is the measured distance from each site to the colony edge. As distance to the prairie dog colony edge increases, the probability of Mountain Plover nest-site use would  decrease, because nests nearer colony edges may give plovers easier access to complementary resources outside colonies: "taller vegetation may provide protection from predators (Schneider et al. 2006), while vegetation and shrubs may provide better opportunities for thermoregulation, especially important for chicks" (Discussion Section Paragraph 4).

## c. Table of models

Make a table of all the models you will need to run. You will run 4 models: a null model, a saturated model, and two other models with different combinations of predictors.

Your table should have 4 columns: (1) model number, (2) visual obstruction, (3) distance to prairie dog colony, and (4) model description.

| model # | visual obs. | distance to prairie dog colony |model_description
|---------|-------------|--------------------------------| --------------------|
| 1       | No          | No                             | null model          |
| 2       | Yes         | Yes                            | saturated model     | 
| 3       | Yes         | No                             | visual obs.         |
| 4       | No          | Yes                            | distance only       | 


## d. Run the models

Create a clean dataset to use called `mopl_clean`. In no particular order:

- clean the column names
- create a new column called `used_bin` to create 1s and 0s from the `id` column
- select `used_bin`, `edgedist`, and `vo_nest`

Write your code to run all your models. Do not display any output.

```{r results=hide}
mopl_clean <- mopl |>
  clean_names() |>  # make column names lowercase and consistent
  mutate(used_bin = if_else(id == "used", 1, 0)) |>
  select(used_bin, edgedist, vo_nest)  # keep only variables needed for modeling
# Model 1: null model
mod_1 <- glm(used_bin ~ 1, 
             data = mopl_clean, 
             family = binomial)

# Model 2: saturated model
mod_2 <- glm(used_bin ~ edgedist + vo_nest, 
             data = mopl_clean, 
             family = binomial)

# Model 3: edgedist only
mod_3 <- glm(used_bin ~ edgedist,
             data = mopl_clean,
             family = binomial)

# Model 4: vo_nest only
mod_4 <- glm(used_bin ~ vo_nest,
             data = mopl_clean,
             family = binomial)
```


## e. Select the best model

Using Akaike’s Information Criterion (AIC), choose the best model.

```{r}
AIC(mod_1, mod_2, mod_3, mod_4)
```

In text, write what the best model was (i.e. “The best model as determined by Akaike’s Information Criterion (AIC)…”).

**Use the predictors and the response variable to describe the model**, not the model number that you assigned.

The best model out of the 4 as determined by Akaike’s Information Criterion (AIC) was the model predicting probability of Mountain Plover nest site using only the variable `vo_nest` (visual obstruction height). This model had the lowest AIC value (331.77) compared with the models indicating that visual obstruction alone provided the best balance between model fit and complexity.

## f. Check the diagnostics
Check your diagnostics for the model you selected using simulated residuals from the `DHARMa` package.

Display all code and output.

```{r}
best_model <- mod_4
simulateResiduals(best_model, plot = TRUE)
```

## g. Visualize the model predictions

Create a plot showing model predictions *for your selected model* with 95% confidence intervals around the predictions and the underlying data.

Show and annotate all code. Show the output.

For full credit:

- make sure the x- and y-axis labels are written in full with units if necessary
- take out the gridlines
- use colors that are different from the default

```{r}
mod_preds <- ggpredict(mod_4, terms = "vo_nest [0:20 by = 1]")

ggplot(mopl_clean,
       aes(x = vo_nest,
           y = used_bin)) +
  geom_jitter(size = 3,
              alpha = 0.4,
              color = "magenta3",
              height = 0.03,
              width = 0) +
  geom_ribbon(data = mod_preds,
              aes(x = x,
                  y = predicted,
                  ymin = conf.low,
                  ymax = conf.high),
              inherit.aes = FALSE,
              alpha = 0.4,
              fill = "cyan2") +
  geom_line(data = mod_preds,
            aes(x = x,
                y = predicted),
            inherit.aes = FALSE,
            linewidth = 1,
            color = "purple4") +
  labs(x = "Visual obstruction height at nest (cm)",
       y = "Probability of Mountain Plover nest site use",
       title = "Predicted probability of Mountain Plover nest site use") +
  scale_y_continuous(limits = c(0, 1),
                     breaks = c(0, 1)) +
  theme_minimal() +
  theme(panel.grid = element_blank())
```

## h. Write a caption for your figure.

Include a figure number, title, description of the figure, and data citation.

**Figure 1. Predicted probability of Mountain Plover nest site use decreases as visual obstruction at the nest increases.** Transparent pink circles (jittered points) represent individual observations of available and used nest sites, plotted as binary outcomes (0 = available, 1 = used) against visual obstruction height at the nest (cm). The dark purple line represents the fitted predicted probability of nest site use from the selected logistic regression model, and the light blue ribbon represents the associated 95% confidence interval around those predictions. Data originally from: Duchardt, Courtney; Beck, Jeffrey; Augustine, David (2020). Data from: Mountain Plover habitat selection and nest survival in relation to weather variability and spatial attributes of Black-tailed Prairie Dog disturbance [Dataset]. Dryad. <https://doi.org/10.5061/dryad.ttdz08kt7>

## i. Calculate model predictions

Calculate the predicted probability of Mountain Plover nest site occupancy at 0 on the index of visual obstruction.

Show and annotate all code. Display the output.

```{r}
ggpredict(mod_4, terms = "vo_nest [0]")
```

If the index of visual obstruction is 0, the predicted probability of Mountain Plover nest site occupancy is 0.73 (95% CI [0.65, 0.80]).

## j. Interpret your results

Write 3-5 sentences summarizing what you found, making references to the figure you made in part h and the predictions you calculated in part j. Your summary should include your interpretation of:

- the predicted probability of nest site occupancy at 0 (no visual obstruction)
- what relationships, if any, exist between visual obstruction and distance to prairie dog colony and probability of occupancy
- the biology behind the trends you found - what explains the relationship between visual obstruction and probability of Mountain Plover nest site occupancy?

The probability of Mountain Plover nest site occupancy was highest when there was no visual obstruction at the nest: at `vo_nest` = 0, the model predicted an occupancy probability of about 0.73 (95% CI [0.65, 0.80]), meaning a site with completely open visibility had a roughly 73% chance of being used. As shown in **Figure 1. Predicted probability of Mountain Plover nest site use decreases as visual obstruction at the nest increases.**, the predicted probability of nest site use decreased sharply as visual obstruction increased, with the fitted line (purple) dropping toward 0 and the 95% confidence ribbon (blue) following the same downward trend. In contrast, distance to prairie dog colony edge was not part of the best-supported model, which suggests it did not explain nest-site occupancy as well as visual obstruction in this dataset. Biologically, this pattern makes sense because Mountain Plovers are birds of open, sparsely vegetated habitat, so low visual obstruction likely improves their ability to detect predators and is more consistent with the short-structure nesting habitat they prefer.

# Problem 4. Affective and exploratory visualizations
## a. Comparing visualizations

Compare and contrast your affective visualization from Homework 3 and the exploratory visualizations you made for Homework 2. In 1-3 sentences each, explain:

- How are the visualizations different from each other in the way you have represented your data?

The exploratory visualizations from Homework 2 focus on clearly displaying the underlying data and statistical relationships using simple and straightforward plots (boxplots and scatterplots). In contrast, the affective visualization from Homework 3 represents the data more abstractly using color, pattern, and opacity to evoke a visual impression rather than emphasizing exact numeric values. This shifts the focus from statistical interpretation to communicating the overall feeling or magnitude of patterns.

- What similarities do you see between all your visualizations?

All of the visualizations I created are based on the same underlying dataset about time slept and time stressed and each attempt to communicate patterns in these variables. They each organize the data visually so that viewers can detect differences or relationships between variables. Despite stylistic differences, they all attempt to represent the same core information/patterns relating to sleep, stress, and sleep efficiency over time/by day of week.

- What patterns (e.g. differences in means/counts/proportions/medians, trends through time, relationships between variables) do you see in each visualization? Are these different between visualizations? If so, why? If not, why not?

The exploratory plots show that time streesed tends to vary across days of the week and that there is some relationship between time slept and time stressed. The affective visualization communicates a similar pattern but in a more abstract way, suggesting variation in stress intensity through color variation rather than exact values. The underlying patterns are the same because the data remain unchanged, but the exploratory plots make the patterns clearer because they retain the quantitative scale while the affective is more emotional in nature.

- What kinds of feedback did you get during week 9 in workshop or from the instructors? How did you implement or try those suggestions? If you tried and kept those suggestions, explain how and why; if not, explain why not.

xxxx need to do 

## b. Designing an analysis
Based on your response variable and one of your potential predictor variables (your focal variable or otherwise), propose one analysis that we talked about this quarter that you could apply to answer your research question(s).

(NOTE: I slightly modified the aim of my personal data project after homework 2. I pivoted my response variable to be time slept and my main predictor to be sleep efficiency with time spent stress remaining as an additional potential predictor.)

Additionally, explain *why* this analysis is appropriate with regard to:

- your question
- your variables and what *types* of variables they are




## c. Check any assumptions and run your analysis
In this section, create two subsections (in the appropriate order) to:

- check any assumptions
- run the analysis you proposed in part b
- (if applicable) calculate an effect size

Show all code and output.

## d. Create a visualization

Create a visualization that reflects the analysis you ran in part c (i.e. if you compared medians, you should create a plot representing medians; if you built a model, you should show predictions and confidence intervals).

For full marks, your plot should:

- have labelled x- and y-axes with units
- use a different theme than the default
- use different colors than the default
- show the underlying data

## e. Write a caption

Write a caption for the figure you created in part d. Include a figure number, title, description of visual components (including but not limited to shapes, colors, lines, etc.).

**Figure 2.**xxx

## f. Write about your results

In 2-4 sentences, write about your results (with appropriate components in the parenthetical statistical summary).

Include any components that would provide additional context to your interpretation (e.g. differences in means) or allow you to quantify the magnitude of effects (e.g. effect sizes).

xxx

## g. Sharing your affective visualization

This is a component you will complete in workshop during week 10. We will be taking attendance that day.

